---
title: "Using an Interactions Model to Analyze the Relationship between Deep Sea Corals, Latitude, and Ocean Depth"
author: "Bailey Jørgensen"
execute:
  warning: false
  message: false
format:
  html:
    code-fold: true
    toc: true
editor_options: 
  chunk_output_type: console
---

Overview
Note: The numbers below only reflect published records. Additional records may be retained by the Program for further review.

Number of records: 30,850
Number of coral records: 24,768
Number of sponge records: 6,082
Records with images: 245
Record type(s): specimen, still image, video observation
Minimum depth (meters): 1
Maximum depth (meters): 6,369
Number of records flagged as <50 meters depth: 1,763
Time frame: -999 to 2022
Data Set Description: This is a subset of the Deep Sea Coral Research and Technology Program’s National Database. The data set contains museum records of deepwater coral and sponge specimens in the Smithsonian Institution, National Museum of Natural History, Invertebrate Zoology Collection. Changes to the original dataset may have been made to conform to the National Database Schema and for quality control purposes. The collection is world-wide in scope, and represents the largest U.S. museum collection of these taxa, including many type specimens. Further information can be found at: http://invertebrates.si.edu/collections.htm


```{r}
#| eval: true
#| echo: false
# Load libraries
rm(list = ls())
library(tidyverse)
library(here)
library(janitor)
library(sf)
library(tmap)
library(terra)
library(stars)
library(spData)
library(spDataLarge)
```

```{r}
#| eval: true
#| echo: false
# Read in data
corals <- read_csv(here("data", "deep_sea_corals_1843_c601_9be4.csv"))
```

```{r}
# Explore data, in order to filter by species and region
head(corals)
```

```{r}
# Filter data to only include columns I want
corals <- corals %>% 
  clean_names(case = "snake") %>% # Change column names to snake_case
  select(scientific_name, phylum, genus, species, individual_count, latitude, longitude, depth_in_meters, country) %>% # Choose columns I want
  filter(phylum %in% c("Cnidaria")) %>% # Filter to exclude sponges and NA row
  sf::st_as_sf(coords = c("longitude", "latitude"),
               crs = st_crs(4326)) %>% 
              filter(st_is_valid(.))
  
corals
```

```{r}
# How many unique scientific names are there?
n_distinct(corals$scientific_name)
```

```{r}
unique(corals$country)
```

```{r}

```


```{r}
world <- spData::world

tm_shape(world) +
  tm_fill(col = "continent",
          palette = c("slategray", "snow3", "slategray2", "slategray3", "slategray4", "lightslategrey", "lightsteelblue4", "snow4"),
          title = "Continents") +
  tm_shape(corals) +
  tm_dots(col = "phylum",
          palette = "pink2",
          size = 0.05,
          border.col = "black",
          title = "Data Point") +
  # tm_compass(type = "4star",
  #            size = .05,
  #            position = c("left", "top")) +
  tm_layout(main.title = "Global Coral Observations",
            title.position = c("center", "top"),
            title.snap.to.legend = FALSE,
            #frame = TRUE, 
            #legend.frame = TRUE,
            legend.outside = TRUE)
```

From this global map, we can see the huge volume to coral observations present in our dataset. In order to get a more focus in my model, I will filter the datset to a specific continent, and run the analysis on that. It could be interesting to, in the future, run the same analysis on each continent, and then cross-compare them. This would reveal interesting insights about different continents and species, but I also think it could be an interesting view into how collecitng bias might affect the results of such an analusis, since we can see that some continents contain more sample points than others. INSERT DATA BIAS SOURCES FROM ETHICS CLASS IN HERE











## Citations

Smithsonian Institution, National Museum of Natural History. Observation date range: 1860 to 2022. Coral or sponge occurrence observations submitted to the NOAA National Database for Deep Sea Corals and Sponges (www.deepseacoraldata.noaa.gov). DSCRTP Dataset ID: NMNH_IZ. Database version: 20241022-1.



